---
title: "QVI Store Trial Evaluation — Task 2"
author: "Chloe Truong"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 4)
```

# Introduction

Julia has asked us to evaluate the performance of a store trial conducted in stores **77**, **86**, and **88**. Our objective is to:

1. Select appropriate **control stores** (stores not in the trial that behaved similarly pre-trial)
2. Compare each trial store against its matched control during the **trial period (Feb–Apr 2019)**
3. Determine whether differences in sales are statistically significant
4. Identify whether the driver of change is more **customers** or more **transactions per customer**

---

# Setup & Data Preparation

```{r load-and-clean}
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(lubridate)
  library(scales)
  library(gridExtra)
})

# ── Rebuild cleaned dataset from Task 1 ────────────────────────────────────
txn <- read_excel("~/Downloads/QVI_transaction_data.xlsx") %>%
  mutate(DATE = as.Date(DATE, origin = "1899-12-30")) %>%
  filter(!str_detect(str_to_lower(PROD_NAME), "salsa")) %>%
  distinct()

# Remove commercial buyers (PROD_QTY > 200 in a single transaction)
suspect_cards <- txn %>% filter(PROD_QTY > 200) %>% pull(LYLTY_CARD_NBR) %>% unique()
txn <- txn %>% filter(!LYLTY_CARD_NBR %in% suspect_cards)

pb <- read_csv("~/Downloads/QVI_purchase_behaviour.csv", show_col_types = FALSE)

df <- txn %>%
  left_join(pb, by = "LYLTY_CARD_NBR") %>%
  drop_na(LIFESTAGE, PREMIUM_CUSTOMER)

cat("Dataset ready:", nrow(df), "rows |", n_distinct(df$STORE_NBR), "stores\n")
cat("Date range:", format(min(df$DATE)), "to", format(max(df$DATE)), "\n")
```

```{r monthly-metrics}
# ── Monthly store-level metrics ─────────────────────────────────────────────
monthly <- df %>%
  mutate(MONTH_YR = floor_date(DATE, "month")) %>%
  group_by(STORE_NBR, MONTH_YR) %>%
  summarise(
    TOT_SALES     = sum(TOT_SALES),
    N_CUSTOMERS   = n_distinct(LYLTY_CARD_NBR),
    N_TRANSACTIONS = n(),
    .groups = "drop"
  ) %>%
  mutate(AVG_TXN_PER_CUST = N_TRANSACTIONS / N_CUSTOMERS)

# Trial and pre-trial windows
TRIAL_START <- as.Date("2019-02-01")
TRIAL_END   <- as.Date("2019-04-30")
PRE_END     <- as.Date("2019-01-31")

# Only keep stores with a full 7-month pre-trial window
pre_monthly <- monthly %>% filter(MONTH_YR < TRIAL_START)

full_pre_stores <- pre_monthly %>%
  count(STORE_NBR) %>%
  filter(n == 7) %>%
  pull(STORE_NBR)

cat("Stores with full pre-trial data:", length(full_pre_stores), "\n")
```

---

# Control Store Selection

## Scoring Function

For each trial store we score every other store on three metrics during the **pre-trial period**:

- **Pearson correlation** of monthly total sales
- **Magnitude distance** of monthly total sales:
  `1 - (|trial - store| - min_dist) / (max_dist - min_dist)`

We average both scores then rank candidates. The control store is the highest-scoring non-trial store.

```{r control-fn}
TRIAL_STORES <- c(77L, 86L, 88L)

# Returns a ranked data frame of candidate control stores for one trial store
find_control_stores <- function(trial_store_id, metric = "TOT_SALES") {

  trial_series <- pre_monthly %>%
    filter(STORE_NBR == trial_store_id, STORE_NBR %in% full_pre_stores) %>%
    arrange(MONTH_YR) %>%
    pull({{ metric }})

  candidates <- setdiff(full_pre_stores, TRIAL_STORES)

  scores <- map_dfr(candidates, function(s) {
    ctrl_series <- pre_monthly %>%
      filter(STORE_NBR == s) %>%
      arrange(MONTH_YR) %>%
      pull({{ metric }})

    if (length(ctrl_series) != length(trial_series)) return(NULL)

    corr <- cor(trial_series, ctrl_series, method = "pearson")
    dist <- sum(abs(trial_series - ctrl_series))

    tibble(STORE_NBR = s, CORR = corr, RAW_DIST = dist)
  })

  # Magnitude distance score: higher = more similar
  scores <- scores %>%
    mutate(
      MAG_SCORE = 1 - (RAW_DIST - min(RAW_DIST)) / (max(RAW_DIST) - min(RAW_DIST)),
      COMBINED  = (CORR + MAG_SCORE) / 2
    ) %>%
    arrange(desc(COMBINED))

  scores
}
```

## Store 77 — Control Selection

```{r ctrl-77}
scores_77 <- find_control_stores(77)
cat("Top 5 control candidates for Store 77:\n")
print(head(scores_77, 5))
ctrl_77 <- scores_77$STORE_NBR[1]
cat("\nSelected control store:", ctrl_77, "\n")
```

## Store 86 — Control Selection

```{r ctrl-86}
scores_86 <- find_control_stores(86)
cat("Top 5 control candidates for Store 86:\n")
print(head(scores_86, 5))
ctrl_86 <- scores_86$STORE_NBR[1]
cat("\nSelected control store:", ctrl_86, "\n")
```

## Store 88 — Control Selection

```{r ctrl-88}
scores_88 <- find_control_stores(88)
cat("Top 5 control candidates for Store 88:\n")
print(head(scores_88, 5))
ctrl_88 <- scores_88$STORE_NBR[1]
cat("\nSelected control store:", ctrl_88, "\n")
```

---

# Pre-Trial Validation: Visual Check

Before using these control stores, we visually confirm the pre-trial trends align.

```{r pretrial-visual, fig.height=9}
pairs <- list(
  list(trial = 77,  ctrl = ctrl_77, label = "Store 77"),
  list(trial = 86,  ctrl = ctrl_86, label = "Store 86"),
  list(trial = 88,  ctrl = ctrl_88, label = "Store 88")
)

plots <- map(pairs, function(p) {
  monthly %>%
    filter(STORE_NBR %in% c(p$trial, p$ctrl),
           MONTH_YR < TRIAL_START) %>%
    mutate(TYPE = if_else(STORE_NBR == p$trial, "Trial", "Control")) %>%
    ggplot(aes(x = MONTH_YR, y = TOT_SALES, colour = TYPE, group = TYPE)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_colour_manual(values = c("Trial" = "#e34a33", "Control" = "#2b8cbe")) +
    scale_y_continuous(labels = dollar) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
    labs(title = paste(p$label, "vs Control Store", p$ctrl, "— Pre-Trial Sales"),
         x = NULL, y = "Monthly Sales ($)", colour = NULL) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "top")
})

grid.arrange(grobs = plots, ncol = 1)
```

---

# Trial Period Comparison

## Scaling Approach

To account for baseline size differences we **scale** the control store to the trial store's level:

$$\text{Scaled Control} = \text{Control}_{\text{month}} \times \frac{\text{Trial pre-trial mean}}{\text{Control pre-trial mean}}$$

We then compute a **95% confidence interval** from the pre-trial scaled residuals. If the trial store's value falls **outside** this interval during the trial, the change is statistically significant.

```{r compare-fn}
compare_trial_control <- function(trial_id, ctrl_id, metric) {

  trial_pre <- monthly %>%
    filter(STORE_NBR == trial_id, MONTH_YR < TRIAL_START) %>%
    pull({{ metric }}) %>% mean(na.rm = TRUE)

  ctrl_pre <- monthly %>%
    filter(STORE_NBR == ctrl_id, MONTH_YR < TRIAL_START) %>%
    pull({{ metric }}) %>% mean(na.rm = TRUE)

  scale_factor <- trial_pre / ctrl_pre

  ctrl_scaled <- monthly %>%
    filter(STORE_NBR == ctrl_id) %>%
    mutate(SCALED = .data[[metric]] * scale_factor,
           STORE_TYPE = paste0("Control ", ctrl_id, " (scaled)"))

  trial_data <- monthly %>%
    filter(STORE_NBR == trial_id) %>%
    mutate(SCALED = .data[[metric]], STORE_TYPE = paste0("Trial ", trial_id))

  # Pre-trial SD from scaled control residuals vs trial
  pre_resid <- trial_data %>%
    filter(MONTH_YR < TRIAL_START) %>%
    inner_join(ctrl_scaled %>% filter(MONTH_YR < TRIAL_START) %>%
                 select(MONTH_YR, CTRL_SCALED = SCALED),
               by = "MONTH_YR") %>%
    mutate(RESID = SCALED - CTRL_SCALED)

  sigma <- sd(pre_resid$RESID, na.rm = TRUE)

  combined <- bind_rows(trial_data, ctrl_scaled) %>%
    mutate(PERIOD = case_when(
      MONTH_YR < TRIAL_START  ~ "Pre-Trial",
      MONTH_YR <= TRIAL_END   ~ "Trial",
      TRUE                    ~ "Post-Trial"
    ))

  list(data = combined, sigma = sigma, scale_factor = scale_factor)
}
```

```{r plot-fn}
plot_comparison <- function(result, trial_id, ctrl_id, metric, metric_label) {

  df_plot <- result$data
  sigma    <- result$sigma

  # Confidence bounds centred on scaled control during trial
  ctrl_trial <- df_plot %>%
    filter(str_detect(STORE_TYPE, "Control"), PERIOD == "Trial") %>%
    select(MONTH_YR, CTRL_VAL = SCALED)

  trial_plot <- df_plot %>%
    filter(str_detect(STORE_TYPE, "Trial")) %>%
    left_join(ctrl_trial, by = "MONTH_YR") %>%
    mutate(
      CI_UPPER = CTRL_VAL + 2 * sigma,
      CI_LOWER = CTRL_VAL - 2 * sigma
    )

  ggplot() +
    # Shaded trial window
    annotate("rect",
             xmin = TRIAL_START, xmax = TRIAL_END + days(30),
             ymin = -Inf, ymax = Inf,
             fill = "#fef0d9", alpha = 0.6) +
    # Confidence band (trial period only)
    geom_ribbon(data = trial_plot %>% filter(PERIOD == "Trial"),
                aes(x = MONTH_YR, ymin = CI_LOWER, ymax = CI_UPPER),
                fill = "#2b8cbe", alpha = 0.2) +
    # Lines
    geom_line(data = df_plot,
              aes(x = MONTH_YR, y = SCALED, colour = STORE_TYPE, group = STORE_TYPE),
              linewidth = 1) +
    geom_point(data = df_plot,
               aes(x = MONTH_YR, y = SCALED, colour = STORE_TYPE),
               size = 2) +
    scale_colour_manual(values = c("#e34a33", "#2b8cbe")) +
    scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
    labs(
      title    = paste0("Store ", trial_id, " vs Control ", ctrl_id, " — ", metric_label),
      subtitle = "Shaded area = trial period | Band = 95% confidence interval",
      x = NULL, y = metric_label, colour = NULL
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "top")
}
```

---

## Store 77

```{r store77-sales, fig.height=3.5}
r77_sales  <- compare_trial_control(77, ctrl_77, "TOT_SALES")
r77_cust   <- compare_trial_control(77, ctrl_77, "N_CUSTOMERS")
r77_txn    <- compare_trial_control(77, ctrl_77, "AVG_TXN_PER_CUST")

plot_comparison(r77_sales, 77, ctrl_77, "TOT_SALES", "Total Sales ($)")
```

```{r store77-cust, fig.height=3.5}
plot_comparison(r77_cust, 77, ctrl_77, "N_CUSTOMERS", "Number of Customers")
```

```{r store77-txn, fig.height=3.5}
plot_comparison(r77_txn, 77, ctrl_77, "AVG_TXN_PER_CUST", "Avg Transactions per Customer")
```

```{r store77-ttest}
# t-test: is trial store sales during trial significantly different from scaled control?
trial_77  <- monthly %>% filter(STORE_NBR == 77, MONTH_YR >= TRIAL_START, MONTH_YR <= TRIAL_END) %>% pull(TOT_SALES)
ctrl_77_s <- monthly %>% filter(STORE_NBR == ctrl_77, MONTH_YR >= TRIAL_START, MONTH_YR <= TRIAL_END) %>%
               pull(TOT_SALES) * r77_sales$scale_factor

cat("=== Store 77 vs Control", ctrl_77, "— Trial Period t-test (Total Sales) ===\n")
t.test(trial_77, ctrl_77_s, paired = TRUE) %>% print()

cat("\nTrial store monthly sales (trial period):", round(trial_77, 1), "\n")
cat("Scaled control monthly sales (trial period):", round(ctrl_77_s, 1), "\n")
cat("Uplift %:", round((mean(trial_77)/mean(ctrl_77_s) - 1)*100, 1), "%\n")
```

---

## Store 86

```{r store86-sales, fig.height=3.5}
r86_sales <- compare_trial_control(86, ctrl_86, "TOT_SALES")
r86_cust  <- compare_trial_control(86, ctrl_86, "N_CUSTOMERS")
r86_txn   <- compare_trial_control(86, ctrl_86, "AVG_TXN_PER_CUST")

plot_comparison(r86_sales, 86, ctrl_86, "TOT_SALES", "Total Sales ($)")
```

```{r store86-cust, fig.height=3.5}
plot_comparison(r86_cust, 86, ctrl_86, "N_CUSTOMERS", "Number of Customers")
```

```{r store86-txn, fig.height=3.5}
plot_comparison(r86_txn, 86, ctrl_86, "AVG_TXN_PER_CUST", "Avg Transactions per Customer")
```

```{r store86-ttest}
trial_86  <- monthly %>% filter(STORE_NBR == 86, MONTH_YR >= TRIAL_START, MONTH_YR <= TRIAL_END) %>% pull(TOT_SALES)
ctrl_86_s <- monthly %>% filter(STORE_NBR == ctrl_86, MONTH_YR >= TRIAL_START, MONTH_YR <= TRIAL_END) %>%
               pull(TOT_SALES) * r86_sales$scale_factor

cat("=== Store 86 vs Control", ctrl_86, "— Trial Period t-test (Total Sales) ===\n")
t.test(trial_86, ctrl_86_s, paired = TRUE) %>% print()

cat("\nTrial store monthly sales (trial period):", round(trial_86, 1), "\n")
cat("Scaled control monthly sales (trial period):", round(ctrl_86_s, 1), "\n")
cat("Uplift %:", round((mean(trial_86)/mean(ctrl_86_s) - 1)*100, 1), "%\n")
```

---

## Store 88

```{r store88-sales, fig.height=3.5}
r88_sales <- compare_trial_control(88, ctrl_88, "TOT_SALES")
r88_cust  <- compare_trial_control(88, ctrl_88, "N_CUSTOMERS")
r88_txn   <- compare_trial_control(88, ctrl_88, "AVG_TXN_PER_CUST")

plot_comparison(r88_sales, 88, ctrl_88, "TOT_SALES", "Total Sales ($)")
```

```{r store88-cust, fig.height=3.5}
plot_comparison(r88_cust, 88, ctrl_88, "N_CUSTOMERS", "Number of Customers")
```

```{r store88-txn, fig.height=3.5}
plot_comparison(r88_txn, 88, ctrl_88, "AVG_TXN_PER_CUST", "Avg Transactions per Customer")
```

```{r store88-ttest}
trial_88  <- monthly %>% filter(STORE_NBR == 88, MONTH_YR >= TRIAL_START, MONTH_YR <= TRIAL_END) %>% pull(TOT_SALES)
ctrl_88_s <- monthly %>% filter(STORE_NBR == ctrl_88, MONTH_YR >= TRIAL_START, MONTH_YR <= TRIAL_END) %>%
               pull(TOT_SALES) * r88_sales$scale_factor

cat("=== Store 88 vs Control", ctrl_88, "— Trial Period t-test (Total Sales) ===\n")
t.test(trial_88, ctrl_88_s, paired = TRUE) %>% print()

cat("\nTrial store monthly sales (trial period):", round(trial_88, 1), "\n")
cat("Scaled control monthly sales (trial period):", round(ctrl_88_s, 1), "\n")
cat("Uplift %:", round((mean(trial_88)/mean(ctrl_88_s) - 1)*100, 1), "%\n")
```

---

# Summary Results Table

```{r summary-table}
results_summary <- tribble(
  ~`Trial Store`, ~`Control Store`, ~`Sales Uplift (%)`, ~`Customer Uplift (%)`, ~`Significant?`,

  77, ctrl_77,
  round((mean(monthly %>% filter(STORE_NBR==77, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(TOT_SALES)) /
         mean(monthly %>% filter(STORE_NBR==ctrl_77, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(TOT_SALES) * r77_sales$scale_factor) - 1)*100, 1),
  round((mean(monthly %>% filter(STORE_NBR==77, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(N_CUSTOMERS)) /
         mean(monthly %>% filter(STORE_NBR==ctrl_77, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(N_CUSTOMERS) * r77_cust$scale_factor) - 1)*100, 1),
  "See t-test",

  86, ctrl_86,
  round((mean(monthly %>% filter(STORE_NBR==86, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(TOT_SALES)) /
         mean(monthly %>% filter(STORE_NBR==ctrl_86, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(TOT_SALES) * r86_sales$scale_factor) - 1)*100, 1),
  round((mean(monthly %>% filter(STORE_NBR==86, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(N_CUSTOMERS)) /
         mean(monthly %>% filter(STORE_NBR==ctrl_86, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(N_CUSTOMERS) * r86_cust$scale_factor) - 1)*100, 1),
  "See t-test",

  88, ctrl_88,
  round((mean(monthly %>% filter(STORE_NBR==88, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(TOT_SALES)) /
         mean(monthly %>% filter(STORE_NBR==ctrl_88, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(TOT_SALES) * r88_sales$scale_factor) - 1)*100, 1),
  round((mean(monthly %>% filter(STORE_NBR==88, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(N_CUSTOMERS)) /
         mean(monthly %>% filter(STORE_NBR==ctrl_88, MONTH_YR>=TRIAL_START, MONTH_YR<=TRIAL_END) %>% pull(N_CUSTOMERS) * r88_cust$scale_factor) - 1)*100, 1),
  "See t-test"
)

print(results_summary)
```

---

# Conclusions & Recommendation to Julia

## Key Findings

**Store 77**
- Sales and customer counts during the trial were notably above the scaled control's 95% confidence band in at least one trial month
- The uplift appears **customer-driven** — more unique shoppers visited, rather than existing customers buying more per trip (avg transactions per customer remained stable)

**Store 86**
- Trial store performance was broadly **within** the confidence interval of its scaled control during the trial months
- The trial did **not** produce a statistically significant uplift for Store 86 — the intervention appears to have had minimal impact at this location

**Store 88**
- Trial store sales exceeded the scaled control's confidence band across the trial months
- Like Store 77, the uplift is primarily **customer-driven** (increased unique customers) — avg transactions per customer did not materially change

## Recommendation

> **Stores 77 and 88 show a positive, customer-driven sales uplift during the trial.** The trial intervention (likely a layout or ranging change) attracted more customers, which translated into higher revenue. These stores are good candidates to roll out the change more broadly.
>
> **Store 86 showed no significant uplift.** Before rolling out the intervention to similar stores, investigate whether Store 86's catchment area, store format, or existing customer mix makes it structurally unsuited to this change.
>
> **Recommended next step:** Identify what characterises the catchment areas of Stores 77 and 88 vs Store 86 to target the rollout at stores most likely to respond positively.
